Type-Safe Modular Parsing

### 2016-12-12 paper/sections/Overview.tex
Done 2.2, rewriting 2.3

\subsection{Chosing the Parsing Technique}\label{subsec:overview-parsing}

@@

\subsection{The Parsing Expression Problem}
\bruno{You are not motivating the problem! You are going straight to
  the solution without pointing out what the problem is first. What
  you need to do is: First show what happens with conventional
  parsers: at some point, if you add extensions the recursive calls
  will be wrong. Then you show (in the next section) the solution: 
 use delegation/open recursion.}

\bruno{I think we need to set up a challenge here, similar to the challange of
  the expression problem. The challange should be like. Build a parser
for a simple expression language, then extend the language with both 
a new language construct and a new operation. This section will show 
how we can do that using traditional parsers, but without modular
type-safety and separate compilation. The remaining sections will show
that the techniques introduced by us, enable us to solve those two challenges.}

@@

\begin{lstlisting}[language=PlainCode]
Expr := Var String | App Expr Expr
\end{lstlisting}
\bruno{if you intend to write a grammar here, then please use grammar
  (BNF) notation, and not some notation invented by you!}

\paragraph{Extensing the Parser}

\paragraph{Problem: Hardcoded Recursive Calls}

@@

The problem of hardcoded recursive calls can be solved using
delegation, which can be encoded with open recursion. The idea is
simple: instead of making direct recursive calls, an additional
argument, that abstracts over the recursive call, is passed to the 
parsers. This technique enables the developement of parsers that 
support OO-style extensibility of ASTs.

@@

\begin{lstlisting}
abstract class Expr
case class Var(x: String) extends Expr
case class App(e1: Expr, e2: Expr) extends Expr

val pVar: PackratParser[Expr] = ident ^^ Var

val pApp: (=> PackratParser[Expr]) => PackratParser[Expr] =
  e => e ~ e ^^ { case e1 ~ e2 => App(e1, e2) }
\end{lstlisting}
\bruno{You are copying and pasting code into the paper. Don't do this:
code will change alot (various edits), and it is all too easy to get
inconsistent code snippets. Instead use the small Ruby tool that we 
have in the paper directory to import source code directly from the
code repository. This will ensure the consistency of the code
throughout the paper writing. Please ask Haoyuan about the tool; he
knows how to use it!}
\bruno{Please polish up code. You probably want to consistently use 
the variable name ``self'' whenever you want to refer to the whole
parser (instead of ``e'' that you use at the moment). Also you should 
use ``Open'' (which you currently call ``Fix'') instead of using the
longwinded ``(=> PackratParser[Expr]) => PackratParser[Expr]''} 
\bruno{Also, for consistency, why isn't ``pVar'' open? I.e. using a
  self reference as well? I know that it is not recursive, but giving
  it a different treatment makes it more confusing for users.}
\bruno{You should use a more OO approach in the example. 
Rather than case classes show an Expr trait with pretty printing.
This will enable you to talk about the extensibility and object
algebras better later on.
}

The abstract syntax is defined by Scala's case classes, and 
we write parsers for the two cases \lstinline{Var} and \lstinline{App}.
The parser \lstinline{pVar}, for parsing variables, is straightforward. It parses a
single identifier, and its type \lstinline{PackratParser[Expr]}
indicates the result is an \lstinline{Expr}.
The parser \lstinline{pApp}, for parsing applications is more
interesting. \lstinline{pApp} takes a parser of
\lstinline{Expr} as an argument, and uses it to parse two consecutive
expressions as an application. This argument \lstinline{e} represents
the parser of \emph{the whole expression}. In this particular case, it
is in fact the combination of \lstinline{pVar} and
\lstinline{pApp}. We can replace it by \lstinline{pVar | pApp} and the
code still works. However, using an abstract argument provides
extensibility to the parser. Because once the abstract syntax is
extended later, the parser of the whole expression will change
correspondingly. Then we have to edit existing code if it is
hard-coded. The parameterization technique used in \lstinline{pApp} is
called \textit{open recursion}, which is discussed in section
\ref{sec:openandparsing}.

@@

\begin{lstlisting}[language=PlainCode]
Expr := Var String | App Expr Expr | Lam String Expr
\end{lstlisting}
\bruno{same comment as before: if this is supposed to be a grammar,
  then use a grammar notation}

>>>>

\subsection{Choosing the Parsing Technique}\label{subsec:overview-parsing}

@@

\subsection{The Parsing Expression Problem}
\huang{done, added the attempt/failure/reason of extending conventional parsers}\bruno{You are not motivating the problem! You are going straight to
  the solution without pointing out what the problem is first. What
  you need to do is: First show what happens with conventional
  parsers: at some point, if you add extensions the recursive calls
  will be wrong. Then you show (in the next section) the solution:
 use delegation/open recursion.}

\huang{partly done, only extend a new language construct in the example, how about leaving extension of new operations in the OA subsection?}\bruno{I think we need to set up a challenge here, similar to the challange of
  the expression problem. The challange should be like. Build a parser
for a simple expression language, then extend the language with both
a new language construct and a new operation. This section will show
how we can do that using traditional parsers, but without modular
type-safety and separate compilation. The remaining sections will show
that the techniques introduced by us, enable us to solve those two challenges.}

@@

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <var>
    \alt <expr> <expr>
    \alt `(' <expr> `)'
\end{grammar}

It is straightforward to model ASTs by inheritance and write corresponding parsers for each case.

\begin{lstlisting}
abstract class Expr
class Var(x: String) extends Expr
class App(e1: Expr, e2: Expr) extends Expr

trait BaseParser {
  lexical.delimiters += ("(", ")")
  val pExpr: PackratParser[Expr] = pVar ||| pApp ||| "(" ~> pExpr <~ ")"
  val pVar: PackratParser[Expr] = ident ^^ { x => new Var(x) }
  val pApp: PackratParser[Expr] =
    pExpr ~ pExpr ^^ { case e1 ~ e2 => new App(e1, e2) }
}
\end{lstlisting}

As the combination of \lstinline{pVar} and \lstinline{pApp}, \lstinline{pExpr} can parse expressions correctly.

\paragraph{Attempt to Extend the Parser} Now consider we are extending the abstract syntax as well as the parsers. We add lambda abstractions as a new case of expressions. The language then becomes the famous \textit{untyped lambda calculus}.

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= ...
    \alt `\\' <var> `.' <expr>
\end{grammar}

The ASTs can be extended trivially as below.

\begin{lstlisting}
class Lam(x: String, e: Expr) extends Expr
\end{lstlisting}

Since we already have the parser for variables and applications, we would like to build the new parser by reusing the old one. So we compose \lstinline{pLam}, which parses lambda abstractions, with the old parser. \lstinline{pExtExpr} is the new function to parse expressions.

\begin{lstlisting}
trait ExtParser extends BaseParser {
  lexical.delimiters += ("\\", ".")
  val pExtExpr: PackratParser[Expr] = pExpr ||| pLam ||| "(" ~> pExtExpr <~ ")"
  val pLam: PackratParser[Expr] =
    "\\" ~> ident ~ ("." ~> pExtExpr) ^^ { case x ~ e => new Lam(x, e) }
}
\end{lstlisting}

Unfortunately, this extended parser does not work properly to parse all possible expressions defined by the new grammar. For example, it fails to parse \lstinline{(\x.x)} \lstinline{(\y.y)} which is obviously valid.

\paragraph{Problem: Hard-coded Recursive Calls} The critical problem of extending parsers directly is that the recursive calls in old parsers are hard-coded. Specifically, \lstinline{pApp} makes two recursive calls to parse two consecutive expressions as an application. Before the extension, we implemented it by using \lstinline{pExpr} for those recursive calls. It worked fine at that time because \lstinline{pExpr} covers all cases of the old grammar. However, after the extension of lambda abstractions, the proper function for such recursive calls should be \lstinline{pExtExpr} instead of \lstinline{pExpr}, because the new function \lstinline{pExtExpr} covers all of the three cases in the new grammar, including the lambda case.

That is the reason why the extended parser fails to parse some valid expressions such as \lstinline{(\x.x)} \lstinline{(\y.y)}. To resolve this issue, we need to let \lstinline{pApp} be aware of the change of grammar. We have some direct solutions:

\begin{itemize}
    \item Keep the old code, rewrite problematic functions like \lstinline{pApp}. \\Separate compilation is obtained but code reuse is lost, because every function may contain such recursive calls.
    \item Modify the old code, correct the recursive calls by proper functions. \\We can avoid code duplication thus get code reuse, but separate compilation is lost because of modification on the old code.
    \item ??? \huang{third one? use dynamic languages?}
\end{itemize}

Similar to the trade-off in the Expression Problem, neither of them satisfies us. The challenge of building semantically modular parsers requires fundamental change of the parser structure. We will introduce our solution in the next sections.

@@

Hard-coded recursive calls prevent our parsers from extensibility.
This problem can be solved using delegation, which can be encoded with \textit{open recursion}. The idea is
simple: instead of making direct recursive calls, an additional
argument, that abstracts over the recursive call, is passed to the
parsers. This technique enables the developement of parsers that
support OO-style extensibility of ASTs.

@@

\begin{lstlisting}
type Open[T] = (=> T) => T

// Expr, Var and App are the same as previous section
trait ExprParser {
  val pVar: Open[PackratParser[Expr]] =
    _ => ident ^^ { x => new Var(x) }
  val pApp: Open[PackratParser[Expr]] =
    self => self ~ self ^^ { case e1 ~ e2 => new App(e1, e2) }
}
\end{lstlisting}

\bruno{You are copying and pasting code into the paper. Don't do this:
code will change alot (various edits), and it is all too easy to get
inconsistent code snippets. Instead use the small Ruby tool that we
have in the paper directory to import source code directly from the
code repository. This will ensure the consistency of the code
throughout the paper writing. Please ask Haoyuan about the tool; he
knows how to use it!}

\huang{done}\bruno{Please polish up code. You probably want to consistently use
the variable name ``self'' whenever you want to refer to the whole
parser (instead of ``e'' that you use at the moment). Also you should
use ``Open'' (which you currently call ``Fix'') instead of using the
longwinded ``(=> PackratParser[Expr]) => PackratParser[Expr]''}

\huang{done}\bruno{Also, for consistency, why isn't ``pVar'' open? I.e. using a
  self reference as well? I know that it is not recursive, but giving
  it a different treatment makes it more confusing for users.}

\huang{done}\bruno{You should use a more OO approach in the example.
Rather than case classes show an Expr trait with pretty printing.
This will enable you to talk about the extensibility and object
algebras better later on.
}

\lstinline{pApp} takes a parser of
\lstinline{Expr} as an argument, and uses it to parse two consecutive
expressions as an application. This argument \lstinline{e} represents
the parser of \emph{the whole expression}. In this particular case, it
is in fact the combination of \lstinline{pVar} and
\lstinline{pApp}. We can replace it by \lstinline{pVar | pApp} and the
code still works. However, using an abstract argument provides
extensibility to the parser. Because once the abstract syntax is
extended later, the parser of the whole expression will change
correspondingly. Then we have to edit existing code if it is
hard-coded. The parameterization technique used in \lstinline{pApp} is
called \textit{open recursion}, which is discussed in section
\ref{sec:openandparsing}.

@@

\setlength{\grammarindent}{5em}
\begin{grammar}
<expr> ::= <var>
    \alt <expr> <expr>
    \alt `\\' <var> `.' <expr>
\end{grammar}
